```{r}
library(phdR2)
library(dplyr)

```


Mr Custos, Mr Opponent, ladies and gentlemen,

Translation studies is a multi-faceted discipline with a large spectrum of
different research topics and methods. My own interests have been, frankly, not
in the absolute mainstream of this field. The focus of my dissertation in
particular and my work as a researcher in general has not been, for instance,
translating as a profession or the sociological aspects of working as an
interpreter, but rather two languages and their differences on the level of
grammar. It is my conviction that although the work that has been done in
studying different phenomena surrounding the work of translators is valuable,
we ought not to forget about the more linguistically oriented work that
concentrates on the actual substance of translations, that is, the languages
involved.

The part of linguistics that is especially relevant for translation studies has
traditionally been called contrastive linguistics, that is the close
comparison of two or more languages, usually with respect to a certain quality
or phenomenon.  Conducting  contrastive research -- as well as linguistics in a
more general sense -- has experienced a rather large shift during the last
couple of  decades. With the advent of large digitalized collections of texts
-- usually referred to as *corpora* -- new possibilities have arisen and with
the available corpora becoming larger and larger in size the nature of these
possibilities has also changed. With regard to these possibilities
and their implications to translation studies  there are two
points I would like to make.

The first point -- which has been true for quite some time -- is that we can
now work in the direction of moving from observations to generalisations. That
is, we can more realistically describe linguistic phenomena on the basis of
real world data, not just by introspection.  The second point is that we can
also study phenomena that cannot be easily presented  with a set of predefined
rules -- in general, phenomena that are usually considered fuzzy or hard to get
a hold on.  One such phenomenon is the theme of my dissertation, the comparison
of word order in languages where it has traditionally been considered free.

Let me give an illustration of how this works by using one of the questions
addressed in the dissertation. Imagine a teacher teaching Finnish
for students whose first language is Russian. A student has written 
a text -- say, the assignment was to write an email -- with the following
beginning [SCREEN]:

>Eilen minä puhuin Kallen kanssa uudesta työpaikastani. Hän oli innoissaan...

Now, the teacher might give a comment about the beginning of the email, saying
something isn't quite right with it. She might say you probably shouldn't put
the  adverbial in the beginning, although she probably cannot really explain
why.  Or it might be the case that the teacher cannot even define what it is
that is not "quite right" with the text. 

Now, with an empirical approach and explicit comparison of two languages we
can, first of all, find means of explaning why a certain construction does not
"sound" right even though it is perfectly grammatical.  Secondly, we are able
to higlight some potential things  you should be looking at when writing in
your weaker language. That is, we can move the teachers and the learners
towards using patterns that are known to be idiomatic. Instead of giving them a
bucket full of lego blocks and instructions on how to attach individual blocks
to one another we can show them examples of what is already known to look good
and work and what is used frequently.

```{r}

groups <- d %>% 
  filter(group %in% c('L1a', 'L5a', 'L5b')) %>% 
  count(lang, group, location3) %>% 
  group_by( lang, group) %>% 
  mutate(fr=100*n/sum(n)) 

 pl <- groups %>% 
   filter(location3=='S1') %>% 
   ungroup() %>% 
   mutate(group=case_when(
                          group == 'L1a' ~ 'eilen',
                          group == 'L5b' ~ 'viime viikolla',
                          group == 'L5a' ~ 'viime vuonna')) %>% 
   mutate(lang=case_when(
                          lang == 'fi' ~ 'Finnish',
                          lang == 'ru' ~ 'Russian')) %>% 
   ggplot(aes(x=lang, y=fr, fill=lang)) +
   geom_bar(stat="identity" ) + 
   geom_label(aes(label=paste0("n=",n))) + 
   facet_grid( ~group)
```

```{r}
pl

```




So, going back to the example on the screen and its connection to my
dissertation. In one part of the thesis I make a comparison of three Finnish
and Russian temporal expressions: viime vuonna, viime viiikolla and eilen. It
turns out that these three have quite different usage patterns in Finnish and
Russian. If we look at the Russian вчера, we can say that of a total of around
4000 Russian SVO sentences it is used in the beginning in about 2500 cases (more
than 60%). Of a total of 3000 Finnnish sentences, on the other hand, only about
500 (less than 20%) have the adverbial in the beginning.  The difference
between the languages is more than 40 percents (on the screen).  I will not go
into the statistical details here. For the purposes of the current
presentation, suffice it to say that the differences get
smalller, if we compare "last week" and still smaller if we compare "last
year.".

So, when we look at sentences in isolation, the interlingual differences tend
to look rather fuzzy and hard to explain, perhaps to the extent that one begins
to doubt wether or not there even is any difference at all. Going back to the
example on the screen, the teacher might not have any good way of describing
why a particular example with a sentence-initial *eilen* doesn't sound quite
right in Finnish or why a Russian student would write such a sentence in the
first place. However, when we take a broader look at the usage of time
adverbials in a large dataset, we can get clues on what kind of patterns
are frequent and what are not. Clearly, the usage of vthsera at the beginning
of the sentence forms a pattern that is not as strongly present in Finnish with
the Finnish *eilen*. On the other hand, since the differences between Finnish
and Russian are smaller when it comes to *last week* and especially
*last year*, we can also discern that the patterns that are more frequent in Russian
are at least somewhat peculiar to vtshera. Going back to the example, thinking about what is not quite
right with the student's sentence, the teacher can at least say, that these
kind of sentences seem to be more typical in Russian than in Finnish and simply
inform the student that this is a fact he should be aware of when writing in Finnish.

These kind of discoveries, however, pose a challenge to the researcher.
Finding empirical differences between two languages is, of course, already a
valuable insight by itself -- an insight that is beneficial from the
perspective of language learners and, specifically, translators. Having
knowledge about the frequency of different usage patterns is the type of skill
that I don't think is discussed enough in translator education and should be
encouraged. Nonetheless, general observations about the existence of these
differences can only get us so far: it is another task to try to
account for *why* it is the case that, for instance, the sentence-initial
vtshera is so much more frequent in Russian than the sentence-initial eilen in
Finnish.

In trying to accomplish this task I have found it useful to make a distinction
between what is sometimes called in the methodological literature direct and
indirect evidence. The word orders of Finnish and Russian mostly serve
pragmatic purposes, which means that by, for instance, positioning a time
expression in location A rather than location B, the speaker or writer
makes a choice about how to package information and in what tone or style
he or she wants to deliver the information to his interlocutor or reader.
The problem is, that these kind of choices are rather difficult to track
down automatically for statistical analysis. About the only way to get *direct*
evidence about the information packaging of a sentence would be to have somebody
manually annotate the data for such notions as topic and focus or the activation
states of different referents.

One of the hypotheses I make in the dissertation concerning the question
about why the sentence-inital vtshera is so frequent in Russian is that
in Russian it is a typical strategy to construct a text by 
starting a whole sequence with a time adverbial. 
I have labelled these kind of usage patterns introductory  
construcions, of which you can see one example on the screen:

> Вчера вторая ракетка Открытого чемпионата Франции Мария Шарапова провела свой первый матч на парижских грунтовых кортах. (RuPress: Советский спорт)

<!--
eilen toinen tennispelaaja avoin-GEN mestaruuskilpailu-GEN Ranska-GEN Marija Šarapova suorittaa-PRET POSS ensimmäinen-ACC ottelu-ACC PREP pariisilaiset-PREP massa-PREP kentät-PREP
\'[eilen] Ranskan avoimen tennisturnauksen toiseksi sijoitettu pelaaja Marija Šarapova pelasi ensimmäisen ottelunsa Pariisin massakentillä.\'
-->

Now if the hypothesis is that these kind of sentences are more typical for
Russian than for Finnish, how could we characterize them? Clearly, an important
feature is that since they are located at the beginning of a new sequence --
usually a whole news story like in this case or some new episode in a narrative
text -- they should contain words that refer to things new to the reader.  Or,
to be more precise, they should contain referents that the reader
is unaware of, that is, not thinking about at the moment of
reading the sentence. So, one could probably test the hypothesis
about the prevalence of the introductory construction by
testing how often a sentence-initial vtshera is accompanied by referents
that reader is not already thinking about.

This is the crucial methodological point to make. How *can* you test such a
thing as wether or not a person is probably thinking about a referent at the
moment of reading a sentence -- especially when trying to analyze large masses
of texts and not isolated examples? As mentioned, one way would be to have the texts
annotated, so that someone would actually have read each sentence and then manually
marked some words as referring to entities a potential reader is not thinking 
about. That  would, however, be not only tedious, time-consuming and difficult, but also
rather unreliable. And this is where the usage of indirect evidence 
can be of great help.

In my thesis I assess the introductory construction hypothesis by looking at
two kinds of indirect evidence concerning the novelty of the referents in a
sentence.  First, I analyze the length of subjects: my conclusion is that if a
subject of a sentence consists of several words (like in the example on the
screen) it is more likely to be referring to entities that have not yet been
presented to the reader, whereas when a subject consists of only one word
(say, if the example would have been just "Sharapova" or even more obviously
"she") it is probably something that has just been mentioned. Secondly, I take a look 
at the presence of certain kinds of verbs in a sentence with the assumption
that words like *publish, announce, declare* and similar give a clue about
something new being presented to the reader.

When looking at Finnish and Russian examples with eilen or vtshera with the
help of this kind of indirect evidence, one can make some interesting
observations that take us further than just the initial finding about there
being a difference in how often these expressions are used sentence-initally.
Namely, it does seem to be the case that having a multi-word subject like in
the example above or a verb like *publish* increases the probability of using
the sentence-initial location in Russian compared to Finnish.  Again, the
actual statistical models for testing these differences can be found in the
thesis, but just as a quick overview let me mention some of the relevant
percentages: 

```{r}

all_locs <- d %>% 
  filter(group == 'L1a') %>% 
  filter(subjtype == 'long') %>% 
  count(location3, lang)%>% 
  group_by(lang) %>% 
  mutate(frlong=n/sum(n)*100) %>% 
  left_join(
   d %>% 
    filter(group == 'L1a') %>% 
    count(location3, lang)%>% 
    group_by(lang) %>% 
    mutate(frall=n/sum(n)*100), by="lang")

just_l1  <- d %>% 
  filter(group == 'L1a', location3=='S1') %>% 
  count(subjtype, lang) %>% 
  group_by(lang) %>% 
  mutate(fr=n/sum(n)*100)

subjplot <- just_l1 %>% 
  ggplot(aes(x=lang,y=fr, fill=subjtype)) + 
  geom_bar(stat="identity")  +
  geom_label(aes(label=paste0("n=",n)),position=position_stack(vjust=0.5))


pverb1plot <- d %>% 
  filter(group == 'L1a', location3=='S1') %>% 
  mutate(hasPverb=case_when(
                  (headverb %in% L1a_data_for_s2s3$publishverbs$ru$verbi |
                   headverb %in% L1a_data_for_s2s3$publishverbs$fi$verbi) ~ "yes",
                  T ~ "no")) %>% 
  count(hasPverb, lang) %>% 
  group_by(lang) %>% 
  mutate(fr=n/sum(n)*100) %>% 
  ggplot(aes(x=lang,y=fr, fill=hasPverb)) + 
  geom_bar(stat="identity")  +
  geom_label(aes(label=paste0("n=",n)),position=position_stack(vjust=0.5))


```


```{r}
subjplot
```

```{r}
pverb1plot
```

My point here is not, however, to give a detailed description of the findings
in my disertation, but rather to demonstrate the usefullness of certain kinds
of empirical approaches in explaining interlingual differences that cannot be
given clear-cut definitions or boundaries. The location of adverbials of time
in Finnish and Russian are a prime example of such a case. Looking at these
kind of phenomena from a birds-eye perspective using large amounts of data and
trying to establish patterns, probabilities and tendencies is something that I
think should become an integral part of training tranlators and interpreters
and other multilingual communicators that are usually dealing with
a specific pair or some specific pairs of langauges. After all, the argument
can be made that linguistic knowledge in general is statistical by nature: 
that is even L1 learners have been shown to be sensitive to frequencies and
storing in their memory not just information about what a certain
word or expression is but also about when and how often it is used.




<!-- 
In coclusion I want to, once again, emphasize, that the contribution
I am  making with my dissertation ...

...The kind of detailed contastive comparison demonstrated here
and in the dissertation 


[STILL SOMETHING!]

-->

So, in summary, what I have wanted to achieve with my dissertation project (and
more generally as a researcher) is to collect information about the differences
between Finnish and Russian. This aim has been quite practical in nature: to
find out concrete use cases that can be used as a reference when thinking about
the different possible locations of time expressions. As a result of the
project I have produced concrete information about how one should change his or
her ways of constructing discourse and ways of thinking about texts when
writing or speaking in another language. This is the kind of information I
myself found valuable when studying and it is my humble wish that the work I
have produced in the form of the dissertation would become a source of this
type of knowledege for anyone else working with the same kind of problems.


Mr Opponent, I now call upon you to present your critical comments on my
dissertation.



